CODECHECK: An open-science initiative for the independent execution of computations underlying research articles

### 2021-01-12 codecheck-paper.tex
restructure dimensions for better readability

The example workflow leaves room for different
implementations of this process. To elaborate on these options, we consider
the following dimensions in a space of possible CODECHECK workflows, as shown in 
Figure~\ref{fig:dimensions}, to represent relevant aspects for the involved
stakeholders. These aspects touch on timing, responsibilities, and 
transparency and will be discussed in the following sections.
Each of these have their own pros and cons and can for the most part be
freely combined.

@@

The point in time when a CODECHECK happens greatly depends on the 
intentions and the ascribed importance.
The earlier a CODECHECK happens in the process, the more the
results can play a role in editorial decisions:
is a paper published, sent back for revisions, or rejected.
Such a \textbf{strict prerequisite} can reduce the workload of reviewers,
and therefore seems out of place late in the review process.
The later in a review process the check happens, the less challenging is it
to allow a bidirectional communication between author and codechecker, e.g.,
because the author might already be notified of the acceptance and may more
freely share materials online.
Even earlier checks, i.e., a CODECHECK of a preprint, may even help to 
improve the workflow itself even before a publisher is involved. The 
possibility for codechecking papers could be part of a preprint server's
policy, or simply be initiated by interested readers and community members.
% TODO add a sentence about the possible effect of just asking for
% code/data early on in the submission process, see https://twitter.com/tsuyomiyakawa/status/1230672758163402752?s=09)

We see a CODECHECK as part of a typical scholarly communication process.
While more radical ideas to change the paradigms
how researchers share their work exist\footnote{For example Octopus, 
\url{https://science-octopus.org/about}, or Hypergraph
\url{https://www.libscie.org/hypergraph}.},
% mimosa is another platform idea, but no real prototype AFAICT:
% https://projects.invisionapp.com/share/E9Z3RKE7W3P#/screens/436978924
% https://docs.google.com/presentation/d/1V_K8hghgnvGfEtW7TdwtowTveZyy4Qs8rMemNyHx_Dg/edit#slide=id.p
there is a value in an evolutionary approach.
Thereby, existing communities can 
transition towards more open practices at their own pace and ensure
no one is left behind unfairly.
When integrating a CODECHECK in existing review and publication processes,
the \emph{turnaround time} is crucial. Depending on when and who conducts
the check, it may be done faster, or add considerably to the overall
time from submission to publication.
In our first checks, we found that a codecheck may generally take from 2-5
hours, with some outliers on the higher end. This time includes writing and
publishing the certificate, but excludes the time for actual computations.
These efforts include considerable time to communicate about the process, 
especially around who publishes which document when, so that proper 
cross-referencing between paper and certificate is ensured via persistent
identifiers.
When integrated into a peer review system, this handling of documents should
become much more effective and reduce organisational efforts.

@@

A CODECHECK may also be conducted in \textbf{parallel} to the scientific review.
This puts less burden on the turnaround time for the check, yet only makes the 
outcomes available for the final consideration of the handling editor.
The check could also be tasked on demand, e.g., only if one of the reviewers
suggests/requests one because previous screenings might not have caught the 
fact that a computational workflow is part of the submission. However, 
soliciting such a "specialist review" is much more undesirable than having
a regular CODECHECK, without any special treatment for some submissions.
A CODECHECK could then be awarded \textbf{equal weight} as the other reviews.

A \textbf{post acceptance} CODECHECK puts the least requirements on the author 
and may simply lead to an award of excellence on top of the acceptance of the
submission. This \textbf{extra merit} is the least strict solution while still
properly  acknowledging the check results,
because the CODECHECK is to be completed 
before the publication of the final paper.
The GIScience group of checks (see \nameref{register}) falls into this
category: the AGILE conference highlights articles where the reproducibility
was successfully reviewed with a badge on the volume and article landing
pages. Similarly, the GIScience papers in collaborations with journals were 
checked while the authors worked on revisions.

Finally, a CODECHECK may also be conducted \textbf{post publication}, though it
would required to update the article and article metadata later to reference 
the check, otherwise readers would presumably not be aware of it. In general,
publishers are hesitant to make such a revision of published articles.
While this has the least impact on the current publishing practices and also
enables open discourse,
it also has the smallest effect on the original submission
and does not acknowledge the importance of reproducible workflows for
ensuring good scientific practice.

@@

The topic of blinding is broadly discussed, not the least in the push towards
open peer review as part of the Open Science movement 
(cf. \cite{ross-hellauer_guidelines_2019}).
Without taking a strong stance on this discussion, the general motivation 
for more  transparency and reproducibility does indeed favour a more open 
review process.
However, anonymity can also serve to protect individuals 
\cite{tennant_limitations_2020}, e.g.,  junior scientists, whereas it might 
be acceptable to require revealing of the names of tenured researchers.
The possible negative effects of a non-anonymous review are greatly reduced
if a CODECHECK is not relevant for the decision to accept or reject, but that
is of course not desirable. Instead, we argue that a CODECHECK is technical
process that should in any case be fixable and not a question of opinion or
faulty approach. It is technically possible for a nonsense workflow to receive
a CODECHECK certificate.
If passing a CODECHECK becomes mandatory, the full transparency may have to be 
revisited as the relations between authors and codecheckers fall under the
same social and acommunity challenges as non-anonymous peer review
(cf.~\cite{everythinghertz123}).
% previous sentenc inspired by https://osf.io/9cftx/
Furthermore, the level of documentation that is needed for third parties
to reproduce a workflow is extremely hard to get right. Too often, this 
uncertainty leads to researchers giving up and not documenting at all.

The technical nature of the check and the challenge of good enough 
documentation is why we see great benefits in a effective bidirectional
means of communication between author and codechecker. Instead of trying to
fix problems or guess the next step, the codechecker can ask the author to 
rework the docs  or the code.
Instead of lengthily describing a small change, the codechecker may simply
provide a fix within the code.
Instead of hiding useful and  instructive files created during the codecheck
(e.g., a machine-readable computing environment specification), the author 
and readers can profit from the software-related expertise of the
codechecker.
While such a communication may be facilitated in an anonymous way by the 
publisher, it most likely only helps to protect the identity of the 
codechecker, because, especially when of a certain size, not trivial, or 
if very specialised, it is close to impossible to fully anonymise a 
software analysis stack.
Therefore, the most effective means with the most desirable result for
the stakeholders is a non-anonymous and collaborative CODECHECK.
The possible contributions by the codechecker may even be integrated into
the code of the workflow and be acknowledged as code commits. This way, 
proper credit is given within the research software development community.
Nevertheless, depending on other needs represented by other dimensions,
the aspect of `Who knows who?' may be the most flexible one.

@@

The options for the person who can conduct a CODECHECK require a not quite
simple matching of a combination of skills and availability. Ideally, the
codechecker has a matching code \emph{and} domain expertise, for example to
check a workflow based on Python code analysing neuroimaging datasets. However,
a well-documented workflow should be executable by any person with a basic
understanding of computers. Naturally, the more prerequisite knowledge the
codechecker has, the less time is needed to understand the goals and 
the mechanics of an analysis. From our experiences, the priority should be
given to matching technical expertise first, as the a lack of knowledge in
setting  up a computing environment with a particular language or tool is 
much more discouraging and the assessment of the outcome, e.g., comparing
created figures with the original, is largely possible without 
understanding the science.
The time and expertise allotted to the check
are the main drivers for the depth of checking, though in general, we 
expect a CODECHECK not to evaluate performance, e.g., does a racing car
have the best apex speed, but to assess operation free of failure, i.e., 
is it a car with 4 wheels and at least one door, and does it move forwards
when accelerating. This depth can be seen independent from the power of
the check (see above), as long as it is communicated clearly to all
stakeholders, most importantly authors and readers.

@@

We consider a \emph{single codechecker} to be enough, unlike peer review where
multiple referees are common. Because of the technical focus, the 
determination if the concrete outputs of author and codechecker, who can
collaborate to solve any issues, actually match should be a very factual 
process and not
require much debate or opinion. Code usually makes systematic and repeatable
mistakes is is thereby more reliable and auditable than processes controlled
by humans \cite{tibav:42484}, e.g., in a laboratory.

Depending on the importance of the CODECHECK
(see \ref{when-to-do-a-codecheck}), a mediator, e.g., the editor, or a
second opinion might be warranted in rare cases.

We see a great opportunity to especially involve
\emph{early-career researchers}
(ECRs) as codecheckers. They arguably have a high interest in learning
about the latest tools and technologies as they are building up their own
expertise and specialisation. The possibility to gain insights into latest
science for ECRs and at the same time increase the appreciation of
reproduction efforts is a sweet spot that CODECHECK shares
with \emph{ReScience X}, a journal devoted to reproduction and replication 
experiments \cite{roesch_new_2020}.
In fact, ECRs are often much more familiar with latest technology and are 
likely to be early adopters as creators, i.e., be authors of CODECHECK-ready manuscripts, too\footnote{Cf. the 77\% of 141 registered reports who were 
submitted to leading Neuroscience journals by ECRs
\cite{chambers_registered_2019}.}
They are also introduced to peer review and 
may transition into the role of scientific reviewer over time.
Not to solve the general problems of peer review, which is largely an
unsupervised process, i.e., rarely are reviewers taught how to do it, but
a junior codechecker may be supported by a senior codechecker, too.
We see a high potential in setting up the new processes for CODECHECKING with
a clear commitment on openness and transparency, independent of the
respective current peer review process (see \nameref{who-knows-who}).

@@

In contrast, for researchers it can be very important to be publicly credited
with their activity as a reviewer (see TODO CHECK REF \nameref{who-knows-who}).
A regular
review may be listed in public databases (e.g., ORCID
\footnote{\href{https://support.orcid.org/hc/en-us/articles/360006971333-Peer-Review}{https://support.orcid.org/hc/en-us/articles/360006971333-Peer-Review}} or commercial offerings such as Publons
\footnote{\href{https://publons.com/}{https://publons.com/}} or 
ReviewerCredits\footnote{\href{https://www.reviewercredits.com/}{https://www.reviewercredits.com/}}).
A codechecker should be listed just the same. The number of scientific reviews
is of course a very coarse indicator for the community contribution by an 
individual, yet it is unclear if a regular reviewer, who in addition
conducts a CODECHECK to the scientific review, should be credited twice.

The codecheckers community\footnote{\href{https://github.com/codecheckers/codecheckers/}{https://github.com/codecheckers/codecheckers/}}
currently has over 20 individuals who signed up as volunteer codecheckers
following talks given on the project in the last 12 months.
That is even though no checks are actively elicited.
We see a high potential in an open shared list of potential codecheckers 
compared to a private, in-house group, as it may better match expertise
and share the workload.
% TODO tell more about the motivations and commitments that we know about?
The overarching community is also a mean to ensure CODECHECK is a viable
option not just for large publishers, independent of their financial model,
but also for independent no-cost open access journals.

>>>>

The example workflow leaves room for different
implementations of this process. To elaborate on these options, we consider
the several dimensions in a space of possible CODECHECK workflows, as shown in 
Figure~\ref{fig:dimensions}, to represent relevant aspects for the involved
stakeholders. These aspects touch on timing, responsibilities, and 
transparency and will be discussed in the following sections.
The options on each dimension have their own pros and cons and can,
for the most part, be freely combined.

@@

The point in time for a CODECHECK and the ascribed importance are closely
connected, so the dimensions \emph{When} and \emph{Imporance} are described
together.
The earlier a CODECHECK happens in the process, the more the
results can play a role in editorial decisions:
is a paper published, sent back for revisions, or rejected.
Even earlier checks, i.e.,~a CODECHECK of a preprint, may even help to 
improve the workflow itself even before a publisher is involved. The 
possibility for codechecking papers could be part of a preprint server's
policy, or simply be initiated by interested readers and community members.

A \textbf{strict prerequisite} for passing a check can reduce the workload
of reviewers, and therefore seems out of place late in the review process.
The later in a review process the check happens, the less challenging is it
to allow a bidirectional communication between author and codechecker, e.g.,
because the author might already be notified of the acceptance and may more
freely share materials online closer to the paper publication date.

@@

A CODECHECK may also be conducted in \textbf{parallel} to the scientific review.
This puts less burden on the turnaround time for the check, yet only makes the 
outcomes available for the final consideration of the handling editor.
The check could also be tasked after suggestion by a reviewer and remove the
need for a pre-review screening for computational workflows. However, 
soliciting such a ``specialist review'' is much less desirable than having
a regular CODECHECK, without any special treatment for some submissions.
In both cases, a CODECHECK could then be awarded \textbf{equal weight} as 
the other reviews for the editor's decision.

A \textbf{post acceptance} CODECHECK has the smallest impact on editorial 
decisions and may simply provide an \textbf{extra merit} on top of the 
acceptance of the submission. 
This  is the least impactful solution while still evaluating all material
and properly acknowledging the check results from the paper, because the 
check can be completed  before the publication of the paper.
The GIScience group of checks (see below) falls into this
category: the AGILE conference highlights articles where the reproducibility
was successfully reviewed with a badge on the volume and article landing
pages. Similarly, the GIScience papers in collaborations with journals were 
checked while the authors worked on revisions.

A CODECHECK may also be conducted \textbf{post publication}, though it
would required to update the article and article metadata later to reference 
the check, otherwise readers would presumably not be aware of it. In general,
publishers are hesitant to make such a revision of published articles.
While this has the least impact on the current publishing practices and also
enables open discourse,
it also has the smallest effect on the original submission
and does not acknowledge the importance of reproducible workflows for
ensuring good scientific practice.

In general, the timing of a CODECHECK needs to be connected with 
established or typical scholarly communication process.
While more radical ideas to change the paradigms
how researchers can share their work much more granually exist\footnote{
For example Octopus, 
\url{https://science-octopus.org/about}, or Hypergraph
\url{https://www.libscie.org/hypergraph}.},
% mimosa is another platform idea, but no real prototype AFAICT:
% https://projects.invisionapp.com/share/E9Z3RKE7W3P#/screens/436978924
% https://docs.google.com/presentation/d/1V_K8hghgnvGfEtW7TdwtowTveZyy4Qs8rMemNyHx_Dg/edit#slide=id.p
there is a value in an evolutionary approach and we do not suggest
to completely remove the paper as a point in time to share work just yet.
Enhancing existing processes with CODECHECKs allows communities to 
transition towards more open practices at their own pace and ensure
no one is left behind unfairly.
When integrating a CODECHECK in existing review and publication processes,
the \emph{turnaround time} is crucial. Depending on when and who conducts
the check, it may be done faster, or add considerably to the overall
time from submission to publication.
In our first checks, we found that a codecheck may generally take from 2-5
hours, with some outliers on the higher end. This time includes writing and
publishing the certificate, but excludes the time for actual computations.
These efforts are comparable to the time needed to peer review a submission,
which concurs with the efforts some volunteer codecheckers are willing to
spend.
However, the time includes a considerable share of communicating about the
process,  especially around who publishes which document when, so that proper
cross-referencing between paper and certificate is ensured via persistent
identifiers.
When integrated into a peer review system, this handling of documents should
become much more effective and reduce organisational overhead.

@@

The topic of anonymity is broadly discussed, not the least in the push towards
open peer review as part of the Open Science movement 
(cf. \cite{ross-hellauer_guidelines_2019}).
Without taking a strong stance on this discussion, the general motivation 
behind CODECHECK
for higher transparency and reproducibility does indeed favour a more open 
review process.
However, anonymity can also serve to protect individuals 
\cite{tennant_limitations_2020}, e.g.~junior scientists, whereas it might 
be acceptable to require revealing of names from tenured researchers.
The possible negative effects of a non-anonymous review are reduced
if a CODECHECK is not relevant for the decision to accept or reject, but that
is of course not desirable. Instead, we argue that a CODECHECK is technical
process that should in any case be fixable and not a question of opinion or
a faulty approach.
It is technically possible for a nonsense workflow to receive
a CODECHECK certificate.
If passing a CODECHECK becomes mandatory, the full transparency may have to be 
revisited as the relations between authors and codecheckers fall under the
same social and acommunity challenges as non-anonymous peer review
(cf.~\cite{everythinghertz123}).
% previous sentenc inspired by https://osf.io/9cftx/

The technical nature of the check and the challenge of good enough 
documentation is why we see great benefits in a effective bidirectional
means of communication between author and codechecker. Instead of trying to
fix problems or guess the next step, the codechecker can ask the author to 
rework the docs  or the update code.
Of course, the codechecker may also provide a fix within the code instead
of lengthily describing a small change if it is quickly done.
Authors and readers can profit from codecheckers' experience and approach,
as the latter may create useful and instructive files, e.g.,~a machine-readable
computing environment specification, during the check.
While communication between author and codechecker may be facilitated in an 
anonymous way by the 
publisher, it most likely only helps to protect the identity of the 
codechecker, because, especially when of a certain size, not trivial, or 
if very specialised, it is close to impossible to fully anonymise a codebase.
Therefore, the most effective means with the most desirable variant for
the stakeholders is a non-anonymous and collaborative CODECHECK.
The contributions by the codechecker may even be integrated into
the code of the workflow and be acknowledged as code commits. This way, 
proper credit is given within the research software development community.
Nevertheless, depending on needs represented by other dimensions,
the aspect of `Who knows who?' may be the most flexible one.

@@

The options for the person who can conduct a CODECHECK require a not quite
simple matching of a combination of skills and availability. Ideally, the
codechecker has a matching code \emph{and} domain expertise, for example to
check a workflow based on Python code analysing neuroimaging datasets. However,
a well-documented workflow should be executable by any person with a basic
understanding of computers. Naturally, the more prerequisite knowledge the
codechecker has, the less time is needed to understand the goals and 
the mechanics of an analysis. From our experiences, the priority should be
given to matching technical expertise first, as the a lack of knowledge in
setting  up a computing environment with a particular language or tool is 
much more discouraging and the assessment of the outcome, e.g.,~comparing
created figures with the original, is largely possible without 
understanding the science.
The time and expertise allotted to the check
are the main drivers for the depth of checking, though in general, we 
expect a CODECHECK not to evaluate performance, e.g.,~does a racing car
have the best apex speed, but to assess operation free of failure, i.e.,~is
it a car with 4 wheels and at least one door, and does it move forwards
when accelerating. This depth can be seen independent from the power of
the check (see above), as long as it is communicated clearly to all
stakeholders, most importantly authors and readers.

@@

We consider a \emph{single codechecker} to be enough, unlike peer review where
multiple referees are common. Because of the technical focus, the 
determination if the concrete outputs of author and codechecker, who can
collaborate to solve any issues, actually match should be a very factual 
process and not
require much debate or opinion. Code usually makes systematic and repeatable
mistakes is is thereby more reliable and auditable than processes controlled
by humans \cite{tibav:42484}, e.g.,~in a laboratory.
Depending on the importance of the CODECHECK
(see above), a mediator, e.g.,~the editor, or a
second opinion might be warranted in rare cases.

We see a great opportunity to especially involve
\emph{early-career researchers}
(ECRs) as codecheckers. They arguably have a high interest in learning
about the latest tools and technologies as they are building up their own
expertise and specialisation. The possibility to gain insights into latest
science for ECRs and at the same time increase the appreciation of
reproduction efforts is a sweet spot that CODECHECK shares
with \emph{ReScience X}, a journal devoted to reproduction and replication 
experiments \cite{roesch_new_2020}.
In fact, ECRs are often much more familiar with latest technology and are 
likely to be early adopters as creators, i.e.,~be authors of CODECHECK-ready manuscripts, too\footnote{Cf. the 77\% of 141 registered reports who were 
submitted to leading Neuroscience journals by ECRs
\cite{chambers_registered_2019}.}
They are also introduced to peer review and 
may transition into the role of scientific reviewer over time.
Not to solve the general problems of peer review, which is largely an
unsupervised process, i.e.,~rarely are reviewers taught how to do it, but
a junior codechecker may be supported by a senior codechecker, too.
We see a high potential in setting up the new processes for CODECHECKING with
a clear commitment on openness and transparency, independent of the
respective current peer review process (see \emph{Openness} dimension).

@@

In contrast, for researchers it can be very important to be publicly credited
with their activity as a reviewer.
A regular
review may be listed in public databases (e.g.,~ORCID
\footnote{\href{https://support.orcid.org/hc/en-us/articles/360006971333-Peer-Review}{https://support.orcid.org/hc/en-us/articles/360006971333-Peer-Review}} or commercial offerings such as Publons
\footnote{\href{https://publons.com/}{https://publons.com/}} or 
ReviewerCredits\footnote{\href{https://www.reviewercredits.com/}{https://www.reviewercredits.com/}}).
A codechecker should be listed just the same. The number of scientific reviews
is of course a very coarse indicator for the community contribution by an 
individual, yet it is unclear if a regular reviewer, who in addition
conducts a CODECHECK to the scientific review, should be credited twice.

The codecheckers community\footnote{\href{https://github.com/codecheckers/codecheckers/}{https://github.com/codecheckers/codecheckers/}}
currently has over 20 individuals who signed up as volunteer codecheckers
following talks given on the project in the last 12 months.
That is even though no checks are actively elicited in collaboration
with journals.
Their motivations, mentioned in the sign-up issues\footnote{
\url{https://github.com/codecheckers/codecheckers/labels/registration}},
are:
supporting reproducible research and open science,
establishing a culture of open and reproducible science and being part
of making it the norm,
improving coding skills,
giving back to the community,
gaining experience in helping scientists with their code,
reducing the amount of bad scientific code and encouraging a culture that
reduces friction in passing on academic codebases,
learning from other people's mistakes,
and curiosity.
We see a benefits in an open shared list of codecheckers
across journals
compared to a private in-house group, as it may better match expertise
and share the workload.
The overarching community is also a mean to ensure CODECHECK is a viable
option not just for large publishers, independent of their financial model,
but also for independent no-cost Open Access journals.
